---
title: "Before we Start"
teaching: 10
exercises: 5
questions:
- "FIX ME"
objectives:
- "FIX ME"
keypoints:
- "FIX ME"

source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("00-")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## This is not an introduction to R

This is a course on how to plot data using the ggplot2 library in R.

If you have no previous experience with R, you will probably not have a 
good experience in this course.

This course assumes a certain level of knowledge about R. We are not going to cover the basics, and you are expected to know how to use the following functionalities in R before starting this course.

Have R and R-studio installed. Alternatively run everything on rstudio.cloud
Know how to assign values to variables
know what a function is, and how we pass input and parameters to it
be familiar with the %>% operator
know the basic verbs from dplyr of the tidyverse: select filter mutate arrange summarise
be familiar with dataframes
know how to install and load packages
comments 
doing math on variables 
Get the concept of vectors 
subsetting vectors and dataframes 
logical tests 
NA 
read in data from a csv/excel

If any of these topics are unfamiliar, we strongly recommend that you either take one of our introductory courses, read up on the curriculum of one of them, or follow one of the many amazing courses you can find online, before taking this course.


Tag et af vores introducerende kurser LINKS før du melder dig til her.




## Getting ready to plot!
Before we do anything else, we need to make sure that we have installed the necessary libraries.

```{r installing libraries, eval=F}
install.packages("tidyverse")
```

After installing the libraries, we load the libraries, so we can use them:

```{r}
library(tidyverse)
```

## Hvilke data arbejder vi med

Vi arbejder med en klassiker. Diamanter.

```{r}
head(diamonds)
```

Datasættet diamonds er indbygget i ggplot2. Det indeholder priser på 53,940 diamanter,
og parametre på dem:

- carat: diamantens vægt i karat (0.200 gram)
- cut Kvaliteten af hvordan diamanten er slebet (Fair, Good, Very Good, Premium, Ideal)
- color. farven - fra D (der er bedst), til J (der er dårligst)
- clarity et mål for hvor klar diamanten er. Kategorisk variabel, "I1" (dårligst), "SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF" (bedst)
- depth total depth percentage (øh? z divideret med middelværdien af x og y)
- table bredden af toppen - relativt til det bredeste punkt
- price. I USD
- x længde i mm
- y bredde i mm
- z dybde i mm

## One introductory element in R

Factors

One concept that we do not cover in our introductory courses is *factors*.

Factors are the R way of handling categorical variables. Categorical variables,
also known as qualitative variables, are variabels that can take on one and only
one, of a limited, usually fixed, number of possible values. 

Classical examples are the roll of a six-sided die, that can only have one of the
possible outcomes of 1, 2, 3, 4, 5 or 6. The role in reproduction for a mammal
can be only female or male. Either your cell phone has power. Or it dont.

These are examples of intrinsic categorical values. 

We might also discretize continous variables, like how much battery is left on a 
cell phone. The actual value might be 55%. But we are recording only levels of 
0%, 25%, 50%, 75% and 100%. And in that case we will probably record the categorical
value as 50%.

With the data we are working with here, the cut of a diamond can be either
Fair, Good, Very Good, Premium and Ideal. 

These are categorical values. The cut of a diamond is either Fair or not-Fair.

Let us take a look at the actual values of the cut variable (of the first 6 
observations) in our dataset:
```{r}
head(diamonds$cut)
```
We see the values. And we also see that these values have *levels*, the six
categorical values that the cut of a diamond can have.

If we try to convert the values of the cut to numerical values, we can see how 
these values are actually stored by R:

```{r}
as.numeric(head(diamonds$cut))
```
R handles categorical values, factors, by assigning an integer to each value.
5 to "ideal", 4 to "Premium", 2 to "Good" etc. These are the values stored.

Associated with these values are *levels*

```{r}
levels(diamonds$cut)
```

And they appear in a specific order. 

The cut of diamonds are actually not only categorical, but also *ordinal*, meaning
that they are not just in a specific order, but that one level is "better" than
another. We see an indication here:

```{r}
head(diamonds$cut)
```
Where the "<" indicates that "Good" is better than "Fair", but worse than 
"Very good".

We can change the order of the levels of a factor using functions from the 
*forcats* library:

```{r}
fct_relevel(diamonds$cut, "Good", "Ideal") %>% 
  head()
```
Now the order is changed, because we have specified that the first two levels 
should be "Good", followed by "Ideal".

The forcats library have many other functions for working with factors. 

Regarding plotting in R, the important thing to know is that the order of things
in our plots are typically controlled by the order of factors, and that we can
change that order to suit our needs.




{% include links.md %}