---
title: "Our first plot"
teaching: 10
exercises: 5
questions:
- "FIX ME"
objectives:
- "FIX ME"
keypoints:
- "FIX ME"

source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## The grammar of graphics

Plotting using ggplot2 is based on "The Grammar of Graphics" 
https://link.springer.com/book/10.1007/0-387-28695-0, a theoretical treatment of how 
to talk about and conceptualize plots by Leland Wilkinson.

We do not need to know or understand all details of this 620 page book. But some
weird naming conventions follows from this.

What we do need to know, is that based on the grammar of graphics, the layered
structure of a plot using ggplot, is build like this:

```{r eval = F}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <SCALE_FUNCTION> +
  <FACET_FUNCTION> +
  <THEME_FUNCTION>
```

We are going to cover each element in the following.

## ggplot in it self

The first thing we need to provide for ggplot is some data. We are working with
the diamond dataset:
```{r}
ggplot(data = diamonds)
```
This in itself produces an extremely boring plot. But it is a plot!. What is 
missing is information on what exactly it is in the dataset we are trying to plot.
How should our data be *mapped* to the area of our plot? Or, what should we 
have on the X-axis, and what should be on the Y-axis?

We provide that information to ggplot using the *mapping* argument to the 
ggplot function:
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price))
```

This specifies that we want the carat-value on the x-axis, and the value, price, 
of the diamond on the y-axis. 

We are not actually seeing any data, because we have not specified the way the
individual datapoints should be plotted. 

We are going to use points, in order to make a classic scatter plot. We do that
by adding a geom_ function to our plot:
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point()
```
Note that in the template shown above, mapping was placed in the <GEOM_FUNCTION>
rather than in the ggplot function. If we place the mapping in the ggplot function
that mapping will be inherited by the functions we add later, unless we specify
something different.

That means that:
```{r eval = F}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point()
```

and
```{r eval = F}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

will yield the same result.


An exercise:
Try to plot "carat" (the weight of the diamond) against "x" (the length of the diamond) 

note that some outliers, probably erroneous data, are discovered. A length of 0
is highly suspicious.
solution
```{r}
diamonds %>% 
  ggplot(aes(carat, x)) +
  geom_point()
```

Congratulations! You have made your first plot in R! (or maybe not)



{% include links.md %}