---
title: "Different types of plots"
teaching: 10
exercises: 5
questions:
- "What other types of plots can we make?"
objectives:
- "FIX ME"
keypoints:
- "FIX ME"

source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# A collection of different types of plots

Scatterplots are very useful, but we often need other types of plots. In this part of the course, we are going to look at some of the more
common types.


## Histograms

Histograms splits all observations of a variable up in a number of
"bins". It counts how many observations are in each bin. Then we plot a column
with a height equivalent to the number of observations for each bin:

```{r}
diamonds %>% 
  ggplot(aes(carat)) +
  geom_histogram()
```

Note that we get a warning from `geom_histogram` that the number of 
bins by default is set to 30. 30 bins will almost never be the correct
number of bins, and we should chose a better value ourself.

```{r}
diamonds %>% 
  ggplot(aes(carat)) +
  geom_histogram(bins = 25)
```
General rules for choosing the correct number of bins exists. We recommend that you experiment with different number of bins to find the
one you like best.


## Barcharts

Not to be confused with histograms, barcharts count the number of 
observations in different groups. Where the scale in histograms is 
continuous, and split into bins, the scale in barcharts is discrete. 

Here we map the color-variable to the x-axis in the barchart. `geom_bar` counts the number of observations itself - we do not need to
provide a count:

```{r eval=T}
diamonds %>% 
  ggplot(aes(color)) +
  geom_bar()
```

## A small excursion

Why are the columns in the barchart above in that order? 

One might guess that they are simply in alphabetical order. 

Not so! Color is a categorical variable. Diamonds either have the color 
"D" (which is the best color), or another color (like "J", which is the worst).

There are no "D.E" colors, they do not exist on a continous range.

This is called "factors" in R.
The data in a factor can take one of several values, called levels. And the
order of these levels are what control the order in the plot.

The order can be either arbitrary, or there can exist an implicit order in the data,
like with the color of the diamonds, where D is the best color, and J is the
worst. These types of ordered categorical data are called ordinal data.

They look like this:
```{r}
diamonds %>% 
  select(cut, color, clarity) %>% 
  str()
```
Note that even though the colour "D" is better than "E", the levels 
of the color factor indicates that "D<E". 

All this just to say: We can control the order of columns in the plot, by
controlling the order of the levels of the categorical value we are plotting:

```{r}
diamonds %>% 
  mutate(color = fct_rev(color)) %>%  
  ggplot(aes(color)) +
  geom_bar()
```

`fct_rev` is a function that reverses the order of a factor. It comes from the
library `forcats` that makes it easier to work with categorical data.



## boxplots

Boxplots are suitable for visualising the distribution of data. 
We can make a boxplot of a single variable in the data - or we 
can make several boxplots in one plot:

```{r}
diamonds %>% ggplot(aes(x = carat, y = cut)) +
  geom_boxplot()
```

Here we have the variable we are making boxplots of, on the x-axis,
and splitting them up in one plot per cut, on the y-axis.


> ## What is a boxplot?
>
> Boxplots are useful for showing different distributions. The fat line in
> the middle of the box is the median, the two ends of the box is first and 
> third quartile, and the two whiskers (or lines) on both sides of the 
> box shows the minimum and maximum values - excluding outliers, defined for
> this purpose as values that lies more that 1.5 times the interquartile 
> range from the box.
>
{: .callout}

## Violinplots

Boxplots are not necessarily the best option for showing distributions.
A good alternative could be violinplots. They show a *density plot* - basically a histogram with infinite bins - for each group, blotted symmetrically around an axis:

```{r echo = F}
diamonds %>% ggplot(aes(carat, y = cut)) +
  geom_violin()
```


> ## exercise
> The geom_ for making violin plots is `geom_violin`
> Look at the help for `geom_violin` and make a violinplot
> with carat on the x-axis, and cut on the y-axis.
> > ## Solution
> >
> > diamonds %>% 
> >   ggplot(aes(carat, y = cut)) +
> >
> >   geom_violin()
> >
> {: .solution}
{: .challenge}



## And many more

ggplot2 is born with a multitude of different plots. A complete list of plots will be very long, and take up all the time for this course.
Take a look at [The R Graph Gallery](https://r-graph-gallery.com/)
or at [Graphs in R](https://kubdatalab.github.io/R-graphs/) (NB a work in progress), where we will collect weird and wonderful plots, when to use them, when not to 
use them. And how to make them.

ggplot2 is written as an extensible package, meaning that developers can create packages making plots that are not included in ggplot2, or introduce more advanced functionality around plots. Two of the more interesting extensions are:

`ggforce` extends ggplot2 with specialised plottypes.

`gganimate` makes it easyish to make animated plots using ggplot2


{% include links.md %}