---
title: "Scaling and coordinates"
teaching: 10
exercises: 5
questions:
- "FIX ME"
objectives:
- "FIX ME"
keypoints:
- "FIX ME"

source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("025-")
library(tidyverse)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



The coordinate system can be changed. If we want to flip the coordinates, we
could interchange the x and y values in the mapping argument.
Or we could add a coordinate function that changes the coordinate system:

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price, color = color)) +
  geom_point() +
  coord_flip()
```
Other coord_ functions exists, the most commonly used are coord_flip and coord_polar:

mhp lagkage

DET HER SKAL DER NOK KIGGES LIDT PÅ... SÅDAN DIDAKTISK NOK TIL AT JEG SELV 
FORSTÅR HVAD DER FOREGÅR...
```{r}
diamonds %>% 
  mutate(color = as.character(color)) %>% 
  filter(color == "G") %>% 
  ggplot(aes(x= color, fill = cut)) +
  geom_bar(position = "stack") +
    coord_polar("y", start=0) 

  

```


```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price, color = color)) +
  geom_point() +
  coord_polar()
```
That might not be that useful working with diamonds.


## Log -scale

TÆNK LIDT OVER DEN... forklaring. Hvorfor logaritmerer vi. 
Måske bedre eksempel end herunder.

```{r}
diamonds %>% 
  ggplot(aes(carat, price)) +
  geom_point() +
  scale_y_log10()
```






{% include links.md %}