---
title: "Whats next?"
teaching: 10
exercises: 5
questions:
- "FIX ME"
objectives:
- "FIX ME"
keypoints:
- "FIX ME"

source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What should I do next?

First of all: If you do not have data you want to visualize 
already. Find some!

[Kaggle](https://www.kaggle.com/datasets) host competitions 
in machine learning. For the use of those competitions, they
give access to a lot of interesting datasets to work with.
With more than 200.000 datasets at time of writing, it can
be a bit overwhelming, so consider looking at the "[Data Visualization](https://www.kaggle.com/datasets?fileType=csv&tags=13208-Data+Visualization)" category. The link provided only shows datasets saved as CSV-files, and has only about 2.500 datasets.

## Play around!

ggplot2 comes with a lot of functionality. This is the
list of build-in geoms in ggplot2:

```{r liste-over-geomer, echo = F, warning=F, message=F}
library(ggplot2)
library(dplyr)
lsf.str("package:ggplot2") %>% grep("^geom_", ., value = TRUE)

```
ggplot2 is also build as an extensible package, making 
it relatively easy to build extensions, that does things
that ggplot2 is not able to on its own. [This page](https://exts.ggplot2.tidyverse.org/) contains a collection of these.


## Get some inspiration!

The website/book [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) is a great resource for tips, tricks and thinking about visualizations, 
especially the [directory of visualizations](https://clauswilke.com/dataviz/directory-of-visualizations.html). Note, however, that the author does not provide examples of the code you need to write to make the plots.

There exist an online challenge, [#30DayChartChallenge](https://30daychartchallenge.org/), that 
challenges you to create a data visualization on a certain topic each day of april. That can be a bit of a mouthful and 
we are not going to participate ourselves. But! The collection 
of visualizations from 2022 is humongous, and a great place
to find ideas.


## Other online courses

[EdX](https://www.edx.org/search?q=ggplot) offers a multitude of interesting courses - here we link to their ggplot-related courses.

Codecademy also offers free courses (and a certificate of completion if you pay) [This course](https://www.codecademy.com/learn/learn-ggplot-2) offers
a more indepth introduction to ggplot2.




{% include links.md %}